Epistemic Semantics in Guarded String Models\thanks{Thanks guys.}

### 2021-01-15 tex/pdraft.tex
a_1 -> a_h, epik -> epikat and prose cleaning

\begin{document}
\maketitle
\begin{abstract}
  Constructive and computable multi-agent epistemic possible worlds models are
  as sets of guarded string models in an epistemic extension of Kleene Algebra
  with Tests (\textsc{Kat}). The account is framed as a formal language Epik
  (Epistemic \textsc{Kat}) for defining such models. The language is implemented
  by translation into the finite state calculus, and alternatively by modeling
  propositions as lazy lists in Haskell. The syntax-semantics interface for a
  fragment of English is defined by a categorial grammar.
\end{abstract}

@@

Enter Epik, which is a systematic way of defining such models, and can be
applied to natural language semantics, specifically, epistemic semantics and
clausal embedding. The fundamental insight is to identify possible worlds with
strings of primitive events, so that propositions are sets of (guarded) strings,
whose regular sets have a rich algebraic characterization~\cite{kozen1997kleene}
and computational model~\cite{kozen1997kleenecompleteness,kozen2001automata}.
Epik leverages these algebrae to define a mathematical model
(Sections~\ref{sec:model} \&~\ref{sec:logic}) and computationally interpret
using finite state machines (Section~\ref{sec:fst}) and lazy lists of strings
(Section~\ref{sec:lazy})).

{\em Related Work.} Epik synthesizes five antecedents.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.  {\em Action models} in dynamic
epistemic semantics introduce the technique of constructing epistemic
models from primitive alternative relations on events, in order to
capture the epistemic consequences of perceptual and communicative
events \citep{baltag1999logic}.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensional semantics} has investigated how to do the
semantics of intensional complementation in a setting where
compositional semantics is expressed in a finite state calculus
\citep{rooth2017finite,collard2018finite}.

@@

\newcommand{\apH}{\ensuremath{a_1}}
\newcommand{\apT}{\ensuremath{a_0}}
\newcommand{\bpH}{\ensuremath{b_1}}
\newcommand{\bpT}{\ensuremath{b_0}}

\newcommand{\atTH}{\ensuremath{a_{01}}}
\newcommand{\atHT}{\ensuremath{a_{10}}}
\newcommand{\btTH}{\ensuremath{b_{01}}}
\newcommand{\btHT}{\ensuremath{b_{10}}}

@@

To create realistic models, contradictory event sequences must be prohibited.
For example, $\apH\atTH$ would require the coin to simultaneously heads-up and
tails-up. To mechanize this, events in Epik come with Boolean pre-conditions and
post-conditions. In the running example, the Boolean variables $h$ (the coin is
heads-up) and $t$ (the coin is tails-up) represent the coin's state. Then, since
Amy can only observe heads when the coin is heads-up, event $\apH$ has the
invariant condition $h$. Similarly, Amy can only turn the coin from tails to
heads if it shows tails, so $\atTH$ has the tails-up precondition $t$ and the
heads-up postcondition $h$.

@@

Thus far, the presentation is closely related to Kleene Algebra with Hypotheses
(\textsc{Ka+H})~\cite{cohen1994hypotheses}, which permits user-specification of
futher equations beyond the axioms of \textsc{Ka}. Specifically, the state
formula $\varphi$ can be thought of as a hypothesis $\varphi \equiv 1$, and an
event formula $\zeta_a$ for event $a$, is simply the hypothesis $a \equiv
\sum_{\varphi,\varphi' \in \llbracket \zeta_a \rrbracket} \varphi;a;\varphi'$.
Rather than rely on general techniques ~\cite{doumane2019kleene,
  hardin2002elimination, kozen1997kleenecompleteness}, we relalize our specific
hypotheses directly.

The key facet that differentiates Epik from \textsc{Ka+H} is the epistemic
modality, which cannot be naively represented using universal
hypotheses\footnote{The existence of this encoding is an interesting future
  topic}. Sections~\ref{sec:model} and~\ref{sec:logic} discuss the
model-theoretic and logical development of the epistemic operators.

\begin{figure}
  \fbox{
    \begin{minipage}{\columnwidth}
  \(\begin{array}{lcl}
  \multicolumn 3 l {\textrm{state formulas}\hfill (a \in \mathsf{B}) \hfill \mbox{}}\\
  \rho,\sigma,\varphi &::=& a \mid0 \mid 1 \mid\rho+\sigma \mid \rho\,\sigma \mid \bar{\rho} \\
  \multicolumn {3} l {\textrm{effect formulas}}\\
  \zeta,\eta & ::= & \rho \mathbin{:} \sigma \mid \zeta+\eta
    \mid \zeta \mathbin{\&} \eta \mid \bar{\zeta}
  \end{array}\)

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, events of the agents semi-privately
looking at the coin, and events of secretly turning the coin.  The
line beginning with {\tt state} enumerates ${\sf{B}}$. The line
beginning with {\tt restrict} gives the state formula.  The lines
beginning with {\tt event} declare events and their effect formulas.

@@

Similarly, Her alternatives for $\atTH$ (she turns the coin over from heads to
tails) are $\atTH$ and $\atHT$, indicating that she doesn't know, \emph{a
  priori}, whether she's turning the coin from $H$ to $T$ or from $T$ to $H$.
She has the same alternatives for the event $\atHT$. Conversely, When Bob
secretly turns the coin over, in event $\btTH$ or $\btHT$, she doesn't know
anything has happened, so her alternative is the ``no-operation'' or
``no-information'' event event $o$. Bob's event relation is symmetric.

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o  h:h + t:t\\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent amy \\
            \hspace{3mm} o  -> o \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o0 + o1 \\
            \hspace{3mm} b01 -> o0 + o1 \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

\caption{Epik program describing a possible-worlds event sequence
  model for two agents with information about one coin, and events of
  the agents semi-privately looking at the coin, and privately turning
  the coin.}
\label{fig:epik-example}
\label{epikfigure}
\end{figure}

This paper focuses on defining a concrete possible worlds model from an Epik
specification. The models are an extension of guarded-string models for Kleene
Algebra with Tests (\textsc{Kat}). This algebraic theory has model classes including
guarded string models, relational models, finite models, and matrix models. Our
definitions and notation follow \cite{kozen2001automata}. We add syntax and
semantics to cover multi-agent epistemic semantics.

Guarded strings over a finite alphabet ${\sf{P}}$ are like ordinary strings, but
with atoms over a set ${\sf{B}}$ alternating with the symbols from ${\sf{P}}$.
In the algebra described by Figure \ref{epikfigure}, ${\sf{P}}$ is the set of
events $\{\apH,\apT,\bpH,\bpT,\atTH,\atHT,\btTH,\btHT,o\}$, and ${\sf{B}}$ is
$\{h,t\}$. As we already saw in \ex{effect}, $\mathcal{A}^\varphi_{\mathsf B}$
is $\{h\bar{t},\bar{h}t\}$, for which we use the shorthand $\{H,T\}$. A guarded
string over ${\sf{P}}$ and ${\sf{B}}$ is a string of events from ${\sf{P}}$,
alternating with atoms over ${\sf{B}}$, and beginning and ending with atoms. In
this construction, $w_1=H$, $w_2=Ha_1H$, $w_3=Ha_1Hb_1H$, and
$w_4=Ha_1Hb_1Ha_{10}Tb_{01}Hb_1H$.

The discussion of \ex{coin2} mentioned building worlds by incrementing worlds with
events.  This is accomplished in guarded string models with fusion product
$\diamond$, a partial operation that combines two guarded strings, subject to the
condition that the atom at the end of the the first argument is identical to the atom
at the start of the second one.  \ex{fusion1} gives some examples.

@@

\hspace{-2mm}\(H\,b_1\,H \; \diamond \; H \,a_1\, H = H\, b_1\, H \,a_1\, H \)

\hspace{-2mm}\(T\,b_{01}\,H \; \diamond \; T \,a_1\, T = \textit{undefined} \)
\end{example}

Rather than individual guarded strings, elements of a guarded string
model for \textsc{Kat} are sets of guarded strings.  In our application, these
elements have the interpretation of propositions, which are sets of possible
worlds.  In a free guarded string model for \textsc{Kat}, any event can be
adjacent to any atom in a guarded string that is an element of the
underlying set for the algebra.  We instead impose the constraints
coming from the state and effect formulas.  \ex{wf1} defines the
well-formed guarded strings determined by an Epik specification.  Condition (i)
says that each atom is consistent with the state constraint, and condition (ii)
says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with the effect constraint on $e_i$.\footnote{
An alternative is to define equations such as
$\bar{\phi}=0$ (from the state formula $\phi$)
and $a_1\,{=}\,ha_{1}h$ (from the effect formula $h\mathbin{:}h$ for event $a_1$),
and construct a quotient algebra from the equivalence relation generated
by these equations.  This results in equating sets of guarded strings in the free algebra that
differ by guarded strings that are ill-formed according to the state and
effect formulas.  In the development in the text, we instead use
a set of guarded strings that are well-formed according to the state
and effect formulas as the representative of the equivalence class.
}

@@

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics.  The set of possible worlds in the Kripke frame
determined by an Epik specification is the set of well-formed guarded strings.
At this point, we could say that any set of worlds is a proposition, so that the
set of propositions is the power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead define a more
restrictive set of propositions corresponding to the regular sets of
strings. This is deferred to the next section.  Certain sets of well-formed
guarded strings have the addional interpretation of event types.  An event-type
is something that can ``happen'' is different worlds. For example, $a_1$ corresponds to the 
event type $\{H\,a_1\,H\}$, and $a_0$ corresponds to the event type $\{T\,a_0\,T\}$. Event
types need not be singletons.

The construction so far defines a set of worlds from an Epik specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a general modal frame
$\tuple{W,R_1,...,R_n,K}$ consisting of a set of worlds, a world-alternative relation
for each agent, and a set $K$ of propositions, where each proposition is a subset of
$W$ \citep{chagrov1997modal}.\footnote{As explained in the next section, $K$ will not
  be the power set of $W$, rather it will consists of the regular subsets of $W$.}

An Epik specification defines an alternative relation on bare events for each
agent $a$, which is notated ${R}_{a}$, and lifted to a relation $\hat{R}_a$ on
worlds. The basic idea is that when a world $w$ is incremented with an event
$e$, in the resulting world $w \diamond e$, epistemic alternatives for agent $a$
are of the form $w' \diamond e'$, where $w'$ is an alternative to for $a$ in
$w$, and $e'$ is and event-alternative to $e$ for $a$.\footnote{In this it is
  important that the event-alternative relation for an agent is constant across
  worlds. We anticipate that the definition given here produces results
  equivalent to what is found in literature on event alternatives in dynamic
  epistemic semantics, though we have not verified this. That literature
  primarily focuses on mapping an epistemic model for a single time and
  situation to another, and uses general first-order models, rather than guarded
  string models. See \citet{baltag1999logic}, \citet{van2007dynamic}, and
  articles in \citet{ditmarsch2015handbook}. This literature is motivated by
  epistemic logic and AI planning, rather than computable possible worlds models
  in natural language semantics. } This needs to be implemented in a way that
takes account of pre- and post-conditions for events. For this, our approach is
to refer the definition of well-formed guarded strings. \ex{raise1} defines an
epistemic alternative relation on worlds from an alternative relation on bare
events.

@@

Consider a scenario like the one from Figure 1, but with an additional agent
Cal. The base world $T b_0 T c_0 T$ is one where the coin is tails, and first
Bob looks at tails, and them Cal looks at tails. The first event $b_0$ has the
alternatives $b_0$ and $b_1$ for Amy, and the second event $c_0$ has the
alternatives $c_0$ and $c_1$ for Amy. This results in four combinations
$b_0c_0$, $b_0c_1$, $b_1c_0$, and $b_1c_1$. But some of these worlds are bogus,
such as $b_0c_1$, which says that Bob peeked at the coin and saw tails, and Cal
peeked at the same coin and impossibly saw heads. These contradictions are
filtered by post- and pre-conditions of events in the alternative world, so that
the set of alernatives for Amy in $T b_0 T c_0 T$ is $\{T b_0 T c_0 T,H b_1 H
c_1 H\}$, with two world-alternatives instead of four.

\begin{figure}
  \fbox{
    \begin{minipage}{\columnwidth}
      \(\begin{array}{rlrl}
      \multicolumn{4}{l}{\text{events}~~e\in \mathsf{P}} \\
      p,q &\multicolumn{3}{l}{::= e\,|\,\sigma\,|\,p+q\,|\,pq\,|\,p^*\,|\,\neg p\,|\, \Diamond_ap} \\
      \Box_a p & \triangleq \neg\Diamond_a\neg p &
      \bullet & \triangleq \sum_{e \in \mathsf P}e \\      
      p \wedge q &  \triangleq \neg(\neg p + \neg q) &
      p \to q & \triangleq \neg p + q
      \end{array} \)
  \end{minipage}}

\caption{The language of Epik terms and key derived
  operators.}
\label{fig:epik-grammar}
\label{termfigure}
\end{figure}

@@

With modalities and propositional negation added, the signature of $n$-agent
epistemic \textsc{Kat} is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1,
  ..., \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one, $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin ends heads-up.

@@

{\em Standard Epistemic Modalities.}
Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$
as in \ex{encodings}.\footnote{Deeper analyisis of the lexical semantics
  of {\em know} requires adding modeling of presupposition
  \citep{collard2018finite}. The grammar fragment in Section 6 does not model
  the presupposition of {\em know}, except as
  an entailment.}

@@

Different types of reasoners (e.g. accurate, inaccurate,
etc) are modeled using the event-alternatives in an
Epik specification.\footnote{See the discussion of modal axioms \textbf{T} and
  \textbf{D} below.}  The agents in Figure 1 do not always have reliable beliefs,
because of the possibility of secret turning.

\newcommand{\epikspec}{\test,\event,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

>>>>

\begin{document}
\maketitle
\begin{abstract}
  Constructive and computable multi-agent epistemic possible worlds models are
  as sets of guarded string models in an epistemic extension of Kleene Algebra
  with Tests (\textsc{Kat}). The account is framed as a formal language
  Epi\textsc{Kat} (Epistemic \textsc{Kat}) for defining such models. The
  language is implemented by translation into the finite state calculus, and
  alternatively by modeling propositions as lazy lists in Haskell. The
  syntax-semantics interface for a fragment of English is defined by a
  categorial grammar.
\end{abstract}

@@

Enter Epi\textsc{Kat}, which is a systematic way of defining such models, and can be
applied to natural language semantics, specifically, epistemic semantics and
clausal embedding. The fundamental insight is to identify possible worlds with
strings of primitive events, so that propositions are sets of (guarded) strings,
whose regular sets have a rich algebraic characterization~\cite{kozen1997kleene}
and computational model~\cite{kozen1997kleenecompleteness,kozen2001automata}.
Epi\textsc{Kat} leverages these algebrae to define a mathematical model
(Sections~\ref{sec:model} \&~\ref{sec:logic}) and computationally interpret
using finite state machines (Section~\ref{sec:fst}) and lazy lists of strings
(Section~\ref{sec:lazy})).

{\em Related Work.} Epi\textsc{Kat} synthesizes five antecedents.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.  {\em Action models} in dynamic
epistemic semantics introduce the technique of constructing epistemic
models from primitive alternative relations on events, in order to
capture the epistemic consequences of perceptual and communicative
events \citep{baltag1999logic}.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensional semantics} has investigated how to do the
semantics of intensional complementation in a setting where
compositional semantics is expressed in a finite state calculus
\citep{rooth2017finite,collard2018finite}.

@@

\newcommand{\nop}{\ensuremath{o}}

\newcommand{\apH}{\ensuremath{a_h}}
\newcommand{\apT}{\ensuremath{a_t}}
\newcommand{\bpH}{\ensuremath{b_h}}
\newcommand{\bpT}{\ensuremath{b_t}}
\newcommand{\cpH}{\ensuremath{c_h}}
\newcommand{\cpT}{\ensuremath{c_t}}

\newcommand{\atTH}{\ensuremath{a_{th}}}
\newcommand{\atHT}{\ensuremath{a_{ht}}}
\newcommand{\btTH}{\ensuremath{b_{th}}}
\newcommand{\btHT}{\ensuremath{b_{ht}}}

@@

To create realistic models, contradictory event sequences must be prohibited.
For example, $\apH\atTH$ would require the coin to simultaneously heads-up and
tails-up. To mechanize this, events in Epi\textsc{Kat} come with Boolean pre-conditions and
post-conditions. In the running example, the Boolean variables $h$ (the coin is
heads-up) and $t$ (the coin is tails-up) represent the coin's state. Then, since
Amy can only observe heads when the coin is heads-up, event $\apH$ has the
invariant condition $h$. Similarly, Amy can only turn the coin from tails to
heads if it shows tails, so $\atTH$ has the tails-up precondition $t$ and the
heads-up postcondition $h$.

@@

Thus far, the presentation is closely related to Kleene Algebra with
Hypotheses~\cite{cohen1994hypotheses}, which permits user-specification of
futher equations beyond the axioms of \textsc{Ka}. Specifically, the state
formula $\varphi$ can be thought of as a hypothesis $\varphi \equiv 1$, and an
event formula $\zeta_a$ for event $a$, is simply the hypothesis $a \equiv
\sum_{\varphi,\varphi' \in \llbracket \zeta_a \rrbracket} \varphi;a;\varphi'$.
Rather than rely on general techniques ~\cite{doumane2019kleene,
  hardin2002elimination, kozen1997kleenecompleteness}, we realize our specific
hypotheses directly.

Epi\textsc{Kat} is first and foremost a multi-agent epistemic logic, with epistemic
operators that cannot naively be represented as
hypothesis\footnote{Characterizing this relationship precisely is a future
  research question.}. Aside from formalizing the metaphysical modality seen so
far, Sections~\ref{sec:model} and~\ref{sec:logic} develop the epistemic
modality.

\begin{figure}
  \fbox{
    \begin{minipage}{0.96\columnwidth}
  \(\begin{array}{lcl}
  \multicolumn 3 l {\textrm{state formulas}\hfill (a \in \mathsf{B}) \hfill \mbox{}}\\
  \rho,\sigma,\varphi &::=& a \mid0 \mid 1 \mid\rho+\sigma \mid \rho\,\sigma \mid \bar{\rho} \\
  \multicolumn {3} l {\textrm{effect formulas}}\\
  \zeta,\eta & ::= & \rho \mathbin{:} \sigma \mid \zeta+\eta
    \mid \zeta \mathbin{\&} \eta \mid \bar{\zeta}
  \end{array}\)

@@

Epi\textsc{Kat} is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epi\textsc{Kat} program that describes a possible worlds model for two agents
with information about one coin, events of the agents semi-privately
looking at the coin, and events of secretly turning the coin.  The
line beginning with {\tt state} enumerates ${\sf{B}}$. The line
beginning with {\tt restrict} gives the state formula.  The lines
beginning with {\tt event} declare events and their effect formulas.

@@

Similarly, Her alternatives for $\atTH$ (she turns the coin over from heads to
tails) are $\atTH$ and $\atHT$, indicating that she doesn't know, \emph{a
  priori}, whether she's turning the coin from $H$ to $T$ or from $T$ to $H$.
She has the same alternatives for the event $\atHT$. Conversely, When Bob
secretly turns the coin over, in event $\btTH$ or $\btHT$, she doesn't know
anything has happened, so her alternative is the ``no-operation'' or
``no-information'' event event $\nop$. Bob's event relation is symmetric.

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o  h:h + t:t\\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent amy \\
            \hspace{3mm} o  -> o \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o \\
            \hspace{3mm} b01 -> o \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

\caption{Epi\textsc{Kat} program describing a possible-worlds event sequence
  model for two agents with information about one coin, and events of
  the agents semi-privately looking at the coin, and privately turning
  the coin.}
\label{fig:epik-example}
\end{figure}

The sequel focuses on defining a concrete possible worlds model from an Epi\textsc{Kat}
specification. The models are an extension of guarded-string models for Kleene
Algebra with Tests (\textsc{Kat}). This algebraic theory has model classes including
guarded string models, relational models, finite models, and matrix models. Our
definitions and notation follow \cite{kozen2001automata}. We add syntax and
semantics to cover multi-agent epistemic semantics.

Guarded strings over a finite alphabet ${\sf{P}}$ are like ordinary strings, but
with atoms over a set ${\sf{B}}$ alternating with the symbols from ${\sf{P}}$.
In the model described by Figure~\ref{fig:epik-example}, ${\sf{P}}$ is the set of
events $\{\apH,\apT,\bpH,\bpT,\atTH,\atHT,\btTH,\btHT,\nop\}$, and ${\sf{B}}$ is
$\{h,t\}$. As we already saw in \ex{effect}, $\mathcal{A}^\varphi_{\mathsf B}$
is $\{h\bar{t},\bar{h}t\}$, for which we use the shorthand $\{H,T\}$. A guarded
string over ${\sf{P}}$ and ${\sf{B}}$ is a string of events from ${\sf{P}}$,
alternating with atoms over ${\sf{B}}$, and beginning and ending with atoms. In
this construction, $w_1=H$, $w_2=H \apH H$, $w_3=H \apH H \bpH H$, and
$w_4=H \apH H \bpH H \atHT T \btTH H \bpH H$.

The discussion of \ex{coin2} mentioned building worlds by incrementing worlds
with events and maintaining the pre-and-post conditions. This is accomplished in
guarded string models with the fusion product ($x \diamond y$), a partial
operation that combines two guarded strings $x$ and $y$, subject to the
condition that the atom at the end of the $x$ is identical to the atom at the
start of $y$. \ex{fusion1} gives some examples.

@@

\hspace{-2mm}\(H\,\bpH\,H \; \diamond \; H \,\apH\, H = H\, \bpH\, H \,\apH\, H \)

\hspace{-2mm}\(T\,\btTH\,H \; \diamond \; T \,\apH\, T = \textit{undefined} \)
\end{example}

Rather than individual guarded strings, elements of a guarded string model for
\textsc{Kat} are sets of guarded strings. In Epi\textsc{Kat}, these elements
have the interpretation of propositions, which are (regular) sets of possible
worlds. In a free guarded string model for \textsc{Kat}, any event can be
adjacent to any atom in a guarded string that is an element of the underlying
set for the algebra. However, this subsumes the valid worlds defined by the
state and effect formulae. \ex{wf1} defines the \emph{well-formed} guarded
strings (valid worlds) determined by an Epi\textsc{Kat} specification. Condition
(i) says that each atom is consistent with the state constraint, and condition
(ii) says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with its effects\footnote{ An alternative is to
  define equations such as $\bar{\phi}=0$ (from the state formula $\phi$) and
  $\apH\,{=}\,h \apH h$ (from the effect formula $h\mathbin{:}h$ for event
  $\apH$), and construct a quotient algebra from the equivalence relation
  generated by these equations. This results in equating sets of guarded strings
  in the free algebra that differ by guarded strings that are ill-formed
  according to the state and effect formulas. In the development in the text, we
  instead use a set of guarded strings that are well-formed according to the
  state and effect formulas as the representative of the equivalence class.}.

@@

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics. The set of possible worlds in the modal frame
determined by an Epi\textsc{Kat} specification is the set of well-formed guarded
strings, and propositions are regular sets of guarded strings, rather than the
power set \citep{montague1975formal,gallin1975intensional}. Certain sets of
well-formed guarded strings have the addional interpretation of event types. An
event-type is something that can ``happen'' in different worlds. For example,
$\apH$ corresponds to the event type $\{H\,\apH\,H\}$, and $\nop$ to the event
type $\{T\,\nop\,T, H\,\nop\,H\}$.

The construction so far defines a set of worlds from an Epi\textsc{Kat} specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a general modal frame
$\tuple{W,R_1,...,R_n,K}$ consisting of a set of worlds, a world-alternative relation
for each agent, and a set $K$ of propositions, where each proposition is a subset of
$W$ \citep{chagrov1997modal}.\footnote{As explained in the next section, $K$ will not
  be the power set of $W$, rather it will consists of the regular subsets of $W$.}

An Epi\textsc{Kat} specification defines an alternative relation on bare events for each
agent $a$, which is notated ${R}_{a}$, and lifted to a relation $\hat{R}_a$ on
worlds. The basic idea is that when a world $w$ is incremented with an event
$e$, in the resulting world $w \diamond e$, epistemic alternatives for agent $a$
are of the form $w' \diamond e'$, where $w'$ is an alternative to for $a$ in
$w$, and $e'$ is and event-alternative to $e$ for $a$.\footnote{In this it is
  important that the event-alternative relation for an agent is constant across
  worlds. We anticipate that the definition given here produces results
  equivalent to what is found in literature on event alternatives in dynamic
  epistemic semantics, though we have not verified this. That literature
  primarily focuses on mapping an epistemic model for a single time and
  situation to another, and uses general first-order models, rather than guarded
  string models. See \citet{baltag1999logic}, \citet{van2007dynamic}, and
  articles in \citet{ditmarsch2015handbook}. This literature is motivated by
  epistemic logic and AI planning, rather than computable possible worlds models
  in natural language semantics. } This needs to be implemented in a way that
takes account of pre- and post-conditions for events. For this, our approach is
to refer the definition of well-formed guarded strings. \ex{raise1} defines an
epistemic alternative relation on worlds from an alternative relation on bare
events.

@@

Consider a scenario like the one from Figure~\ref{fig:epik-example}, augmented
with an additional agent Cal. The base world $T\,\bpT\,T\,\cpT\,T$ is one where
the coin is tails, and first Bob looks at tails, and them Cal looks at tails.
The first event $\bpT$ has the alternatives $\bpT$ and $\bpH$ for Amy, and the
second event $\bpT$ has the alternatives $\cpT$ and $\cpH$ for Amy. This results
in four combinations $\bpT\cpT$, $\bpT\cpH$, $\bpH\cpT$, and $\bpH\cpH$, some of
which, like $\bpT\cpH$, are bogus. These contradictions are filtered by post-
and pre-conditions of events in the alternative world, so that the set of
alternatives for Amy in $T\,\bpT\,T\,\cpT\,T$ is $\{T\,\bpT\,T\,\cpT\,T,
H\,\bpH\,H\,\cpH\,H\}$, with two world-alternatives instead of four.

\begin{figure}
  \fbox{
    \begin{minipage}{0.96\columnwidth}
      \(\begin{array}{rlrl}
      \multicolumn{4}{l}{\text{events}~~e\in \mathsf{P}} \\
      p,q &\multicolumn{3}{l}{::= e\,|\,\sigma\,|\,p+q\,|\,pq\,|\,p^*\,|\,\neg p\,|\, \Diamond_ap} \\
      \Box_a p & \triangleq \neg\Diamond_a\neg p &
      \bullet & \triangleq \sum_{e \in \mathsf P}e \\      
      p \wedge q &  \triangleq \neg(\neg p + \neg q) &
      p \to q & \triangleq \neg p + q
      \end{array} \)
  \end{minipage}}

\caption{The language of Epi\textsc{Kat} terms and key derived
  operators.}
\label{fig:epik-grammar}
\label{termfigure}
\end{figure}

@@

With modalities and propositional negation added, the signature of $n$-agent
Epi\textsc{Kat} is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1,
  ..., \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the propositional semantic values of English sentences.

Consider the examples in \ex{interpretation1}. The first term, $\bullet^*\,h$ uses
the syntactic sugar $\bullet$, defined in Figure \ref{fig:epik-grammar} to be
the disjunction of the primitive events. Since a world is a well-formed sequence
of events, $\bullet^*$ is the set of worlds. Multiplying by the state symbol $h$
in the term $\bullet^* h$ has the effect of conjoining $h$ with the atom at the
end of the world. So $\bullet^* h$ is the set of worlds where the coin ends
heads-up.

@@

{\em Standard Epistemic Modalities.} Using the modal primitive $\Diamond_a$, and the
dual encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$ as
\ex{encodings}.\footnote{Deeper analysis of the lexical semantics of {\em know}
  requires adding modeling of presupposition \citep{collard2018finite}. The
  grammar fragment in Section 6 does not model the presupposition of {\em know},
  except as an entailment.}

@@

Different types of reasoners (e.g. accurate, inaccurate,
etc) are modeled using the event-alternatives in an
Epi\textsc{Kat} specification.\footnote{See the discussion of modal axioms \textbf{T} and
  \textbf{D} below.}  The agents in Figure 1 do not always have reliable beliefs,
because of the possibility of secret turning.

\newcommand{\epikspec}{\test,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

{\em Guarded String Interpretation.} A term $p$ of the logical language is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$, where superscript captures dependence
on an Epi\textsc{Kat} specification.  Figure \ref{fig:interp2} defines the
interpretation.  The interpretation $\sem{ 1
}^{\epikspec}$ of the multiplicative identity 1 is the set of atoms that
satisfy the state constraint $\varphi$.  Where $b$ is a primitive
Boolean, $\sem{ b }^{\epikspec}$ is the set of atoms that satisfy the
state constraint and where $b$ is true.  Where $e$ is a primitive
event, $\sem{ e }^{\epikspec}$ is the set of guarded strings that have
the form of $e$ flanked by compatible atoms, as determined by the
event formula $\zeta_e$. The product $pq$ is
interpreted with fusion product raised to sets of guarded strings.
Kleene star is interpreted as the union of exponents ($p^n$ is
the $n$-times product of $p$ with itself, with $p^0 = 1$).
Propositional complement is complement relative to the set of worlds.
The epistemic formula $\Diamond_a p$ is interpreted with Kripke
semantics for epistemic modality, as the pre-image of $p$ 
under the world-alternative relation $\hat{R}_a$.
